### 2025-08-01 data-visualization.qmd
20250801 - ggplot2 syntax

It is also important to use color schemes with distinguishable colors.
Good color schemes for sequential, diverging, and qualitative (i.e., categorical) data are provided by ColorBrewer (<https://colorbrewer2.org>) and are available using the `scale_color_brewer()` and `scale_fill_brewer()` functions of the `ggplot2` package [@R-ggplot2], as demonstrated in @fig-barPlotColorScheme (@sec-barPlotModifiedColorScheme).
There are a variety of resources for color schemes that are accessible to color-blind viewers:

@@

@@

ggplot2::ggplot(
  data = rb_seasonal,
  aes(
    x = age,
    y = rushing_yards)) +
  geom_point() +
  geom_smooth() +
  coord_cartesian(
    xlim = c(20,40),
    ylim = c(0,NA),
    expand = FALSE) +
  scale_x_continuous(
    breaks = seq(from = 20, to = 40, by = 5)
  ) +
  scale_y_continuous(
    breaks = seq(from = 0, to = 2500, by = 250)
  ) +
  labs( # add plot labels
    x = "Running Back's Age (years)",
    y = "Rushing Yards (Season)",
    title = "NFL Rushing Yards (Season) by Player Age",
    subtitle = "(Among Running Backs)"
  )
```

>>>>

It is also important to use color schemes with distinguishable colors.
Good color schemes for sequential, diverging, and qualitative (i.e., categorical) data are provided by ColorBrewer (<https://colorbrewer2.org>) and are available using the `scale_color_brewer()` and `scale_fill_brewer()` functions of the `ggplot2` package [@R-ggplot2], as demonstrated in @fig-barPlotColorScheme (@sec-barPlotModifiedColorScheme).
Good colorblind-friendly color schemes for binned (`_b`), continuous (`_c`), and discrete (i.e., categorical; `_d`) scales are available using the `scale_color_viridis_X()` and `scale_fill_viridis_X()` functions of the `ggplot2` package [@R-ggplot2] (where `X` is `b`, `c`, or `d`, depending on whether you want a binned, continuous, discrete scale). 
There are a variety of resources for color schemes that are accessible to color-blind viewers:

@@

## `ggplot2` Syntax {#sec-ggplot2Syntax}

### General Syntax {#sec-ggplot2SyntaxGeneral}

The `ggplot2` package [@R-ggplot2] is based on the grammar of graphics [@ggplot3rdEd], which is the principle that you can build any graph from three components: 1) the data set, 2) the coordinate system, and 3) the geoms—visual marks that represent data points.
The general syntax for plots created using the `ggplot2` package [@R-ggplot2] is:

ggplot2::ggplot(
  data = INSERT, # 1) data set
  mapping = aes(
    x = variableName1, # name of variable on x-axis
    y = variableName2) # name of variable on y-axis
) +
  coord_cartesian( # 2) coordinate system
    xlim = c(NA, NA), # set limits (min and max) of x-axis
    ylim = c(NA, NA)) + # set limits (min and max) of y-axis
  INSERT_GEOM_FUNCTIONS # 3) geom(s)
```

The `coord_cartesian()` function can be used to specify the axis limits; however, this function is not required and the plot will apply sensible limits by default.
In addition, there are optional settings you can specify or modify, such as [axis titles](#sec-scatterplotPlotLabels), [axis labels](#sec-scatterplotAxes), [faceting](#sec-faceting), and the [theme](#sec-scatterplotTheme).

Here is a template including these potential modifications:

ggplot2::ggplot(
  data = INSERT, # 1) data set
  mapping = aes(
    x = variableName1, # name of variable on x-axis
    y = variableName2) # name of variable on y-axis
) +
  coord_cartesian( # 2) coordinate system
    xlim = c(NA, NA), # set limits (min and max) of x-axis
    ylim = c(NA, NA)) + # set limits (min and max) of y-axis
  INSERT_GEOM_FUNCTIONS + # 3) geom(s)
  scale_x_continuous(
    breaks = INSERT # specify x-axis labels
  ) +
  scale_y_continuous(
    breaks = INSERT # specify y-axis labels
  ) +
  labs( # add plot labels
    x = "INSERT", # x-axis title
    y = "INSERT", # y-axis title
    title = "INSERT", # plot title
    subtitle = "INSERT", # plot subtitle
    caption = "Add a caption below plot", # plot caption
    alt = "Add alt text to the plot" # plot alt-text
  ) +
  INSERT_THEME_NAME + # apply theme
  facet_wrap(vars(INSERT_VARIABLES_TO_FACET)) # facet by one or more variables
```

### Geoms {#sec-ggplot2SyntaxGeoms}

Here is a list of common geoms available in `ggplot2` (there are many others):

- [histogram](#sec-histogram): `geom_histogram()`
- [density plot](#sec-densityPlot): `geom_density()`
- [box-and-whisker plot](#sec-boxAndWhiskerPlot): `geom_boxplot()`
- [violin plot](#sec-violinPlot): `geom_violin()`
- [bar plot](#sec-barPlot): `geom_bar()`
- [scatterplot](#sec-scatterplot): `geom_point()`
- [rug plot](#sec-scatterplotRugPlot): `geom_rug()`
- [line chart](#sec-lineChart): `geom_line()`
- statistical summary (e.g., [best-fit line](#sec-scatterplotBestFitLine)): `geom_smooth()`

### Color {#sec-ggplot2SyntaxColor}

In `ggplot2`, you can specify the color of a geom using one of two parameters: 1) `fill` and 2) `color`.
The `fill` parameter specifies the color inside the geom.
The `color` parameter specifies the outline/border color of the geom.

The colors to use in the `fill` and `color` parameters can be specified by name (as in @fig-colorByName), hexadecimal (HEX) codes (as in @fig-colorByHEX), RGB values (as in @fig-colorByRGB), or by palettes (e.g., ColorBrewer or viridis; as in Figures [-@fig-multipleColorsByPaletteColorBrewer] and [-@fig-multipleColorsByPaletteViridis]).

```{r}
#| label: fig-colorByName
#| fig-cap: "Specifying the Color/Fill by Name."
#| fig-alt: "Specifying the Color/Fill by Name."

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(position_group %in% c("QB","RB","WR","TE")),
  mapping = aes(
    x = fantasyPoints,
    fill = position_group)
) +
  geom_histogram(
    color = "black",
    fill = "forestgreen"
  )
```

You can see a full list of color names available using the `colors()` function:

```{r}
colors()
```

However, you can use any color by specifying its HEX code, as in @fig-colorByHEX.

```{r}
#| label: fig-colorByHEX
#| fig-cap: "Specifying the Color/Fill by Hexadecimal Code."
#| fig-alt: "Specifying the Color/Fill by Hexadecimal Code."

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(position_group %in% c("QB","RB","WR","TE")),
  mapping = aes(
    x = fantasyPoints,
    fill = position_group)
) +
  geom_histogram(
    color = "#000000",
    fill = "#228B22"
  )
```

you can use any color by specifying its values in red-green-blue (RGB) color space using the `rgb()` function, as in @fig-colorByRGB.

```{r}
#| label: fig-colorByRGB
#| fig-cap: "Specifying the Color/Fill by RGB Value."
#| fig-alt: "Specifying the Color/Fill by RGB Value."

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(position_group %in% c("QB","RB","WR","TE")),
  mapping = aes(
    x = fantasyPoints)
) +
  geom_histogram(
    color = rgb(
      red = 0,
      green = 0,
      blue = 0,
      maxColorValue = 255
    ),
    fill = rgb(
      red = 34,
      green = 139,
      blue = 34,
      maxColorValue = 255
    )
  )
```

We can allow the colors to differ by group by specifying the `fill` parameter in the mapping aesthetics, as in @fig-colorByGroup.

```{r}
#| label: fig-colorByGroup
#| fig-cap: "Specifying Color By Group."
#| fig-alt: "Specifying Color By Group."

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(position_group %in% c("QB","RB","WR","TE")),
  mapping = aes(
    x = fantasyPoints,
    fill = position_group)
) +
  geom_density(
    color = "black",
    alpha = 0.7
  )
```

We can specify multiple colors manually using the `scale_fill_manual()` function, as in @fig-multipleColorsManually.

```{r}
#| label: fig-multipleColorsManually
#| fig-cap: "Specifying Multiple Colors Manually."
#| fig-alt: "Specifying Multiple Colors Manually."

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(position_group %in% c("QB","RB","WR","TE")),
  mapping = aes(
    x = fantasyPoints,
    fill = position_group)
) +
  geom_density(
    color = "black",
    alpha = 0.7
  ) +
  scale_fill_manual(
    values = c(
      "QB" = "red", # or could use hex, rgb, etc.
      "RB" = "blue", # or could use hex, rgb, etc.
      "WR" = "green", # or could use hex, rgb, etc.
      "TE" = "yellow") # or could use hex, rgb, etc.
    )
```

We can also specify the colors using a palette, such as by the ColorBrewer palette using the `scale_fill_brewer()` function (as in @fig-multipleColorsByPaletteColorBrewer) or the viridis palette using the `scale_fill_viridis_d()` function (as in @fig-multipleColorsByPaletteViridis).

```{r}
#| label: fig-multipleColorsByPaletteColorBrewer
#| fig-cap: "Specifying Multiple Colors Using the Color Brewer Palette."
#| fig-alt: "Specifying Multiple Colors Using the Color Brewer Palette."

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(position_group %in% c("QB","RB","WR","TE")),
  mapping = aes(
    x = fantasyPoints,
    fill = position_group)
) +
  geom_density(
    color = "black",
    alpha = 0.7
  ) +
  scale_fill_brewer(palette = "Set1") # see ?scale_fill_brewer for list of available palettes
```

```{r}
#| label: fig-multipleColorsByPaletteViridis
#| fig-cap: "Specifying Multiple Colors Using the Viridis Palette."
#| fig-alt: "Specifying Multiple Colors Using the Viridis Palette."

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(position_group %in% c("QB","RB","WR","TE")),
  mapping = aes(
    x = fantasyPoints,
    fill = position_group)
) +
  geom_density(
    color = "black",
    alpha = 0.7
  ) +
  scale_fill_viridis_d(option = "viridis") # see ?scale_fill_viridis_d for list of available palettes
```

### Line Type {#sec-ggplot2SyntaxLineType}

Common line types include "solid" lines, "dashed" lines, and "dotted" lines.
A full list of line types available is provided here: <https://ggplot2.tidyverse.org/articles/ggplot2-specs.html> (archived at <https://perma.cc/N9ED-6LUS>).
To specify the line type, you can use the `linetype` parameter, as in @fig-lineType.

```{r}
#| label: fig-lineType
#| fig-cap: "Specifying Line Type."
#| fig-alt: "Specifying Line Type."

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(player_display_name %in% c("Tom Brady")),
  mapping = aes(
    x = season,
    y = fantasyPoints,
    group = player_display_name)
) +
  geom_line(
    linetype = "dashed"
  )
```

We can allow the line type to differ by group by specifying the `linetype` parameter in the mapping aesthetics, as in @fig-lineTypeByGroup.

```{r}
#| label: fig-lineTypeByGroup
#| fig-cap: "Specifying Line Type By Group."
#| fig-alt: "Specifying Line Type By Group."

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(player_display_name %in% c("Tom Brady","Peyton Manning")),
  mapping = aes(
    x = season,
    y = fantasyPoints,
    linetype = player_display_name)
) +
  geom_line()
```

We can specify the line types manually using the `scale_linetype_manual()` function as in @fig-lineTypeByGroupManually.

```{r}
#| label: fig-lineTypeByGroupManually
#| fig-cap: "Specifying Line Type By Group Manually."
#| fig-alt: "Specifying Line Type By Group Manually."

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(player_display_name %in% c("Tom Brady","Peyton Manning")),
  mapping = aes(
    x = season,
    y = fantasyPoints,
    group = player_display_name,
    linetype = player_display_name)
) +
  geom_line() +
  scale_linetype_manual(
    values = c(
      "Tom Brady" = "solid",
      "Peyton Manning" = "dashed")
    )
```

### Line Width {#sec-ggplot2SyntaxLineType}

To specify the line width, you can use the `linewidth` parameter, as in @fig-linewidth.

```{r}
#| label: fig-linewidth
#| fig-cap: "Specifying Line Width."
#| fig-alt: "Specifying Line Width."

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(player_display_name %in% c("Tom Brady")),
  mapping = aes(
    x = season,
    y = fantasyPoints,
    group = player_display_name)
) +
  geom_line(
    linewidth = 4
  )
```

### Point Shape {#sec-ggplot2SyntaxPointShape}

Common point shapes include circles, triangles, squares, etc.
A full list of point shapes available is provided here: <https://ggplot2.tidyverse.org/articles/ggplot2-specs.html> (archived at <https://perma.cc/N9ED-6LUS>).
To specify the point shape, you can use the `shape` parameter, as in @fig-pointShape.

```{r}
#| label: fig-pointShape
#| fig-cap: "Specifying Point Shape."
#| fig-alt: "Specifying Point Shape."

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(player_display_name %in% c("Tom Brady")),
  mapping = aes(
    x = season,
    y = fantasyPoints,
    group = player_display_name)
) +
  geom_point(
    shape = 2
  )
```

We can allow the point shape to differ by group by specifying the `shape` parameter in the mapping aesthetics, as in @fig-pointShapeByGroup.

```{r}
#| label: fig-pointShapeByGroup
#| fig-cap: "Specifying Point Shape By Group."
#| fig-alt: "Specifying Point Shape By Group."

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(player_display_name %in% c("Tom Brady","Peyton Manning")),
  mapping = aes(
    x = season,
    y = fantasyPoints,
    shape = player_display_name)
) +
  geom_point()
```

We can specify the point shape manually using the `scale_shape_manual()` function as in @fig-pointShapeByGroupManually.

```{r}
#| label: fig-pointShapeByGroupManually
#| fig-cap: "Specifying Point Shape By Group Manually."
#| fig-alt: "Specifying Point Shape By Group Manually."

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(player_display_name %in% c("Tom Brady","Peyton Manning")),
  mapping = aes(
    x = season,
    y = fantasyPoints,
    shape = player_display_name)
) +
  geom_point() +
  scale_shape_manual(
    values = c(
      "Tom Brady" = 1,
      "Peyton Manning" = 3)
    )
```

### Size of Points and Text {#sec-ggplot2SyntaxSize}

To specify the size of points and text, you can use the `size` and `base_size` parameters, respectively, as in @fig-sizePointsText.

```{r}
#| label: fig-sizePointsText
#| fig-cap: "Specifying Size of Points and Text."
#| fig-alt: "Specifying Size of Points and Text."

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(player_display_name %in% c("Tom Brady")),
  mapping = aes(
    x = season,
    y = fantasyPoints,
    group = player_display_name)
) +
  geom_point(
    size = 2 # size of points
  ) +
  theme_classic(
    base_size = 20 # font size
  )
```

You can also specify the font size separately for particular elements, as in @fig-sizeTextManually.

```{r}
#| label: fig-sizeTextManually
#| fig-cap: "Specifying Size of Text Manually for Particular Elements."
#| fig-alt: "Specifying Size of Text Manually for Particular Elements."

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(player_display_name %in% c("Tom Brady")),
  mapping = aes(
    x = season,
    y = fantasyPoints,
    group = player_display_name)
) +
  geom_point() +
  theme(
    axis.title.x = element_text(size = 30), # font size for x-axis title
    axis.title.y = element_text(size = 30), # font size for y-axis title
    axis.text.x  = element_text(size = 12), # font size for x-axis labels
    axis.text.y  = element_text(size = 12)  # font size for y-axis labels
  )
```

@@

ggplot2::ggplot(
  data = rb_seasonal,
  aes(
    x = age,
    y = rushing_yards)) +
  geom_point() +
  geom_smooth() +
  coord_cartesian(
    xlim = c(20,40),
    ylim = c(0,NA),
    expand = FALSE) +
  scale_x_continuous(
    breaks = seq(from = 20, to = 40, by = 5)
  ) +
  scale_y_continuous(
    breaks = seq(from = 0, to = 2500, by = 250)
  ) +
  labs( # add plot labels
    x = "Running Back's Age (years)", # x-axis title
    y = "Rushing Yards (Season)", # y-axis title
    title = "NFL Rushing Yards (Season) by Player Age", # plot title
    subtitle = "(Among Running Backs)" # plot subtitle
  )
```
